Modelling fine-scale dive behavior with hidden Markov models

### 2020-06-25 Report/Sections/appendix.tex
Update Report

@@

A one-dimensional state-switching Ornstein-Uhlenbeck process $y^*$ is the solution to the following stochastic differential equation:
%
$$dy^*_t = \beta_{x^*_t}(\gamma_{x^*_t} - y^*_t)dt + \omega_{x^*_t} dW_t$$
%
where $x^*_t$ is the fine-scale behavior of the animal at time $t$, $\beta_{x^*_t}$ relates to rate at which the process returns to its mean value, $\gamma_{x^*_t}$ is the long-term mean value of the process, $\omega_{x^*_t}$ is related to short-term variance, and $W$ is a Wiener process. As before, $x^*_t$ is described by an unobserved Markov process. The solution to this equation is known to be the following \cite{Michelot:2019}:
\begin{align*}
y^*_{t+\delta} \sim \mathcal{N}\left((1-e^{-\beta_{x^*_t}\delta})\gamma_{x^*_t} + e^{-\beta_{x^*_t}\delta} y^*_t,\quad \frac{\omega_{x^*_t}^2}{2\beta_{x^*_t}} (1-e^{-2\beta_{x^*_t}\delta})\right)
\end{align*}
Now, suppose that $\delta$ is constant for all observations, as is the case for hidden Markov models. In addition, introduce the following transformations:
\begin{align*}
\mu_{x^*_t} = \gamma_{x^*_t}, \qquad \phi_{x^*_t} = e^{-\beta_{x^*_t}\delta}, \qquad \sigma^2_{x^*_t} = \frac{\omega_{x^*_t}^2}{2\beta_{x^*_t}} (1-e^{-2\beta_{x^*_t}\delta})
\end{align*}
Then, we have the following:
\begin{align*}
y^*_{t+\delta} \sim \mathcal{N}\left((1-\phi_{x^*_t})\mu_{x^*_t} + \phi_{x^*_t} y^*_t,\quad \sigma_{x^*_t}^2 \right)
\end{align*}
%
If $\delta$ is fixed and $x^*_t$ is adjusted to follow a Markov chain rather than a Markov process, then this model is equivalent to the CarHMM with normal emission probabilities. Note that all of the parameter transformations above are one-to-one, so it is easy to go from the CarHMM to the continuous model and back again. This allows for the pricipled construction of the continuous-time model to be combined with the computational convienence of the CarHMM.

%\subsection{Proof of correlation between $d_{t-1}$ and $d_t$}

%THIS IS WRONG
%The correlation between $d_{t-1}$ and $d_t$ is defined as $\left(\mathbbm{E}(d_t d_{t-1}) - \mu_{RL,b}^2 \right) / \sigma_b^2$. Let the probability distribution function of $d_t$ and $d_{t-1}$ be denoted as $p$. Then:

%\begin{align*}
%	\mathbbm{E}(d_t d_{t-1}) &= \int_0^\infty \int_0^\infty x_t x_{t-1} p(x_{t-1}) p(x_t|x_{t-1}) dx_t dx_{t-1} \\
	%
%	&= \int_0^\infty x_{t-1} p(x_{t-1}) \left( \int_0^\infty x_t p(x_t|x_{t-1}) dx_t \right) dx_{t-1} \\
	%
%	&= \int_0^\infty x_{t-1} p(x_{t-1}) \left( (1-\phi_b)\mu_{RL,b} + \phi_b x_{t-1} \right) dx_{t-1} \\
%	&= (1-\phi_b)\mu_{RL,b} \int_0^\infty x_{t-1} p(x_{t-1}) dx_{t-1} + \phi_b \int_0^\infty x_{t-1}^2 p(x_{t-1}) dx_{t-1} \\
%	&= (1-\phi_b)\mu_{RL,b}^2 + \phi_b (\mu_{RL,b}^2 + \sigma_b^2) \\
%	&= \mu_{RL,b}^2 + \phi_b\sigma_b^2
%\end{align*}

%So:

%\begin{align*}
%	corr(d_t,d_{t-1}) &= \frac{\mathbbm{E}(d_t d_{t-1}) - \mu_{RL,b}^2}{\sigma_b^2} \\
%	&= \frac{\mu_{RL,b}^2 + \phi_b\sigma_b^2 - \mu_{RL,b}^2}{\sigma_b^2} \\
%	&= \phi_b
%\end{align*}

>>>>

\subsection{Detailed Description of Data Simulation}

$\hat{Y}^*_{t,s^*}$ was simulated using the following procedure. Note that the $k^{th}$ Fourier mode of $\hat{Y}^*_{t,s^*}$ is denoted as $\hat{Y}^{*(k)}_{t,s^*}$:

\begin{align*}
	(\hat{Y}^{*(0)}_{t,0}|X^*_{t,0} = i) &\sim \mathcal{N} \left(0, \sigma^{*(i)} \right) & \\
	%
	(\hat{Y}^{*(0)}_{t,s^*}|X^*_{t,s^*} = i) &\sim \mathcal{N} \left(\phi^{*(i)} * \hat{Y}^{*(0)}_{t,s^*-1}, \sigma^{*(i)} \right), & s^* = 1,2,\ldots, \lfloor Y_t/2 \rfloor \\
	%
	\hat{Y}^{*(k)}_{t,s^*} &= a_{t,s^*}^{(k)} i\sqrt{b^{(k)}_{t,s^*}}, & k = 1,\ldots,49 \\
\end{align*}
\begin{align*}
    a_{t,s^*}^{(k)} &\sim  \left\{\begin{array}{lr}
	-1 & \quad w.p. \enspace 1/2 \\
	1  & \quad w.p. \enspace 1/2
	\end{array}\right. \\
	%
	(b^{(k)}_{t,s^*}|X^*_{t,s^*}  = 1) &\sim {\rm{Gamma}}(1/k^2, 1) \\
	%
	(b^{(k)}_{t,s^*}|X^*_{t,s^*} = 2) &\sim \left\{\begin{array}{lr}
	{\rm{Gamma}}(1/k^2, 1), \quad & k \notin \{1,2\} \\
	{\rm{Gamma}}(100,1), & k = 1 \\
	{\rm{Gamma}}(50,1), & k = 2
	\end{array}\right. 
\end{align*}

\begin{align*}
    \hat{Y}^{*(50)}_{t,s^*} &= 0 & \\
	%
	\hat{Y}^{*(k)}_{t,s^*}  &= -\hat{Y}^{*(100-n)}_{t,s^*}, & \qquad k = 51,\ldots,99
\end{align*}

$Y^*_{t,(100s^*+1):(100s^*+100)}$ was set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,s^*}$:

$$Y^*_{t,(100s^*+1):(100s^*+100)} = IDFT\left(\hat{Y}^*_{t,s^*}\right), \qquad s^* = 0,\ldots,\lfloor Y_t/2 \rfloor - 1$$

There are several practical reasons behind this construction. First, $\hat{Y}^*_{t,s^*}$ is anti-symmetric about $\hat{Y}^{*(50)}_{t,s^*}$ so that its inverse Fourier transform is real-valued. $\hat{Y}^{*(k)}_{t,s^*}$ also decays like $1/k^2$ so that $Y^*_{t,t^*}$ remains continuous within a 2-second window. Note, however, that $Y^*_{t,s^*}$ is not continuous \textit{between} windows, as can be seen in (fig \ref{fig:fourier_example}). However, the jumps are not too severe since $\hat{Y}^{*(0)}_{t,s^*}$ and $\hat{Y}^{*(0)}_{t,s^*+1}$ are highly correlated. See (fig. \ref{fig:sim_data}) for details.

From here it is straightforward to calculate both $Z^{*(1)}$ and $Z^{*(2)}$ after setting $\tilde{f}$. We pick $\tilde{f} = 10$ periods per window, or 5 hertz. Note that $Z^{*(1)}_{t,s^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,s^*}\right)$, so selecting $\sigma^{*(i)}$ and $\phi^{*(1)}$ appropriately will ensure the distributions from (eqn \ref{}).

Note that $Z^{*(2)}$, is the sum of gamma distributions with the same scale parameter, so the distribution of $Z^{*(2)}$ is also a Gamma distribution:

$$Z^{*(2)} = \sum_{k=1}^{10} b^{(k)}_{t,s^*}$$
$$Z^{*(2)} \sim {\rm{Gamma}}

\iffalse

@@

A one-dimensional state-switching Ornstein-Uhlenbeck process $y^*$ is the solution to the following stochastic differential equation:
%
$$dy^*_t = \beta_{x^*_t}(\gamma_{x^*_t} - y^*_t)dt + \omega_{x^*_t} dW_t$$
%
where $x^*_t$ is the fine-scale behavior of the animal at time $t$, $\beta_{x^*_t}$ relates to rate at which the process returns to its mean value, $\gamma_{x^*_t}$ is the long-term mean value of the process, $\omega_{x^*_t}$ is related to short-term variance, and $W$ is a Wiener process. As before, $x^*_t$ is described by an unobserved Markov process. The solution to this equation is known to be the following \cite{Michelot:2019}:
\begin{align*}
y^*_{t+\delta} \sim \mathcal{N}\left((1-e^{-\beta_{x^*_t}\delta})\gamma_{x^*_t} + e^{-\beta_{x^*_t}\delta} y^*_t,\quad \frac{\omega_{x^*_t}^2}{2\beta_{x^*_t}} (1-e^{-2\beta_{x^*_t}\delta})\right)
\end{align*}
Now, suppose that $\delta$ is constant for all observations, as is the case for hidden Markov models. In addition, introduce the following transformations:
\begin{align*}
\mu_{x^*_t} = \gamma_{x^*_t}, \qquad \phi_{x^*_t} = e^{-\beta_{x^*_t}\delta}, \qquad \sigma^2_{x^*_t} = \frac{\omega_{x^*_t}^2}{2\beta_{x^*_t}} (1-e^{-2\beta_{x^*_t}\delta})
\end{align*}
Then, we have the following:
\begin{align*}
y^*_{t+\delta} \sim \mathcal{N}\left((1-\phi_{x^*_t})\mu_{x^*_t} + \phi_{x^*_t} y^*_t,\quad \sigma_{x^*_t}^2 \right)
\end{align*}
%
If $\delta$ is fixed and $x^*_t$ is adjusted to follow a Markov chain rather than a Markov process, then this model is equivalent to the CarHMM with normal emission probabilities. Note that all of the parameter transformations above are one-to-one, so it is easy to go from the CarHMM to the continuous model and back again. This allows for the principled construction of the continuous-time model to be combined with the computational convenience of the CarHMM.

\fi

### 2020-06-25 Report/Sections/background.tex
Update Report

Hidden Markov models are useful when inferring a single unobserved process, but biological processes often involve multiple simultaneous hidden processes which can occur and at different time scales. For example, a preliminary observation of the killer whale dive data shown in figure (\ref{fig:data}) shows that the behavior of this killer whale changes between approximately hour-long periods of predominately short, shallow dives and long, deep dives. Leos-Barajas et al. encounter a similar issue when modeling the movement of a harbor porpoise in the North Sea, and use it as a motivating example when they introduce hierarchical hidden Markov models.

\begin{figure}[h!]
	\centering
	\includegraphics[width=5in]{../Plots/raw_data.png}
	\caption{Raw depth data of a Killer Whale off the coast of British Columbia, Canada.}
	\label{fig:data}
\end{figure}

@@

\textit{Hidden Markov models} (or HMMs) are comprised of a sequence of unobserved states, $X_t$ $(t = 1, \ldots, T)$ which form a Markov chain and a sequence of (possibly high-dimensional) observations $Y_t$ $(t = 1, \ldots, T)$ which are often referred to as ``emissions". The index $t$ often represents equi-spaced jumps in time. In the field of animal movement, the unobserved chain usually represents the latent behaviour of an animal (e.g. foraging, resting, migrating, etc.), while the observations are often a series of step lengths and turning angles for land animals and either depth or accelerometer data (or both) for marine animals. Each random variable in the unobserved chain $X_t$ can take one of $N$ possible values, and $X$ has corresponding initial distribution $\delta \in \bbR^N$ and probability transition matrix $\Gamma \in \bbR^{N \times N}$:
%
$$\delta_i = Pr(X_1 = i); \quad i = 1,\ldots,N$$
%
$$\Gamma_{ij} = Pr(X_{t+1} = j | X_t = i); \quad t = 1, \ldots, T-1; \enspace i,j = 1,\ldots,N $$
%
We assume that $\delta$ is the stationary distribution of the $\Gamma$, implying that the hidden process runs infinitely before observations are taken. Further, the distribution of each observation $Y_t$ depends only on the value of $X_t$ and none of the preceding observations or behavioral states, i.e. $p(y_t|x_1,\ldots, x_N, y_1,\ldots,y_{t-1},y_{t+1},\ldots,y_N) = p(y_t|x_t)$. In particular, we denote the distribution of $Y_t$ given $X_t = i$ as $f^{(i)}(y_t)$. $f^{(1)},\ldots,f^{(N)}$ in turn depend upon a vector of parameters $\Theta = (\theta^{(1)},\ldots,\theta^{(N)})$ and hereafter are referred to as ``emission distributions". A visualization of the dependence structure can be seen in figure (\ref{fig:HMM}).
In order to estimate the parameters of an HMM, suppose data $y = (y_1,\ldots,y_T)$ is observed as a realization of an unknown HMM model. The probability transition matrix $\Gamma$, and parameters of the emission distributions $\Theta$ can be estimated by maximizing the likelihood of $y$, which we denote as $\calL_{\text{HMM}}(y;\Theta,\Gamma)$, with respect to $\Gamma$, $\delta$, and $\Theta$. In addition, $\calL_{\text{HMM}}(y;\theta,\Gamma,\delta)$ can be calculated using the \textit{forward algorithm} \cite{Zucchini:2016} as follows:
%
$$\calL_{\text{HMM}}(y;\theta,\Gamma) = \delta P(y_1;\theta) \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}$$
%
where $\mathbf{1}_N$ is an $N$-dimensional column vector of ones, and:
%
$$P(y_t;\theta) = \text{diag}(f^{(1)}(y_t),\ldots , f^{(N)}(y_t)).$$
%
To remove constraints when maximizing $\calL_{\text{HMM}}(y)$, we parameterize $\Gamma$ using $\eta \in \bbR^{N \times N}$ and the following link function \cite{Barajas:2017}:
%
$$\Gamma_{ij} = \frac{\exp(\eta_{ij})}{\sum_{j'=1}^N \exp(\eta_{i(j')})}, \qquad \eta_{ii} = 0 \quad \forall i \in \{1, \ldots, N\}.$$
%
$\eta_{ii}$ is set to zero for identifiability. $\calL_{\text{HMM}}(y;\theta,\Gamma)$ can be maximized using any numerical optimizer.

@@

A hierarchical hidden Markov model (or HHMM) is a variation of a hidden Markov model in which $X_t$ satisfies the conditions of section \ref{subsection:HMMs}, but in addition it produces yet another fine-scale hidden Markov model of length $T_t^*$. This fine-scale hidden Markov model can be either instead of \cite{Barajas:2017} or in addition to \cite{Adam:2019} a coarse-scale observation $Y_t$ which must satisfy the conditions of section \ref{subsection:HMMs}. The fine-scale HMM is comprised of a fine-scale Markov chain $X^*_t = (X^*_{t,1}, \ldots, X^*_{t,T^*_t})$ and fine-scale observations $Y^*_t = (Y^*_{t,1}, \ldots, Y^*_{t,T^*_t})$, each of length $T^*_t$. $X^*_t$ can take one of $N^*$ values. Note that while $N^*$ can be set to depend upon $X_t$, we assume here that it does not for simplicity in both the model and the notation. Conditioned on the coarse-scale hidden state $X_t = i$, $X^*_t$ is characterized by an initial distribution $\delta^{*(i)} \in \bbR^{N^*}$ and probability transition matrix $\Gamma^{*(i)} \in \bbR^{N^* \times N^*}$:
%
$$\delta^{*(i)}_{i^*} = Pr(X^*_{t,1} = i^* | X_t = i); \quad i^* = 1,\ldots,N^*$$
%
$$\Gamma^{*(i)}_{i^*j^*} = Pr(X^*_{t,t^*+1} = j^* | X^*_{t,t^*} = i^*, X_t = i); \quad t^* = 1, \ldots, T^*_t-1; \enspace i^*,j^* = 1,\ldots,N^*$$
%
Again, although not necessary, we assume that the initial distribution $\delta^{*(i)}$ is the stationary distribution of $\Gamma^{*(i)}$. In total, there are $N$ different values that $\Gamma^{*(i)}$ and $\delta^{*(i)}$ can take- one for each possible value of $X_t$. We denote the sets of all fine-scale initial distributions and probability transition matrices as $\delta^*$ and $\Gamma^*$, respectively, where $|\delta^*| = |\Gamma^*| = N$.

Each fine-scale observation $Y^*_{t,t^*}$ depends only on the value of its corresponding fine-scale hidden state $X^*_{t,t^*}$ and coarse-scale hidden state, $X_t$. In particular, $p(y^*_{t,t^*} | y^*_{t,1}, \ldots y^*_{t,t^*-1},y^*_{t,t^*+1},\ldots,y^*_{t,T^*_t},x^*_t,x,y) = p(y^*_{t,t^*}|x^*_{t,t^*},x_t)$. We denote the distribution of $Y^*_{t,t^*}$ given that $X_t = i$ and $X^*_{t,t^*} = i^*$ as $f^{*(i,i^*)}(y^*_{t,t^*})$. Conditioned on the fact that $X_t = i$, the fine-scale distributions $f^{*(i,1)}, \ldots, f^{*(i,N^*)}$ are parameterized by an $N^*$-dimensional vector of parameters $\Theta^{*(i)} = (\theta^{*(i,1)},\ldots,\theta^{*(i,N^*)})$, where $\theta^{*(i,i^*)}$ describes the distribution $f^{*(i,i^*)}$. Depending upon the behaviour and responses being modelled, it is possible to force the fine-scale emission distributions to be equal across all values of $i \in \{1, \ldots, N\}: \enspace \theta^{*(1,i^*)} = \theta^{*(2,i^*)} = \ldots = \theta^{*(N,i^*)}$. A visualization of the full structure of the HHMM can be seen in figure (\ref{fig:HHMM}).

@@

One of the key assumptions of both HMMs and HHMMs is \textit{conditional independence} between observations at both the coarse and fine scale. Namely, given the state $X_t$, $Y_t$ is assumed to be independent from all other observations. Therefore, traditional HMMs and HHMMs can fail when the observations $Y$ exhibit significant correlation in time. Examples include fluking in marine mammals in Vancouver, BC (see the results section) and the swimming behavior of horn sharks off the coast of Southern California \cite{Adam:2019}.

One way to deal with auto-correlation in fine-scale behavioral processes is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \cite{Lawler:2019}. In particular, the CarHMM describes the distribution of $(Y_t|Y_{t-1}, X_t=s)$ with parameters $\theta^{(s)} = \{\mu^{(s)},\sigma^{2(s)},\phi^{(s)}\}$, where:
%
$$\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=i) = \phi^{(i)}*y_{t-1} + (1-\phi^{(i)}) * \mu^{(i)}$$
$$\mathbb{V}(Y_t| X_t = i) = \sigma^{2(i)}$$
%
The CarHMM therefore explicitly models auto-correlation into the emission distributions of the HMM while maintaining the structure needed to run the forward algorithm and only adding one additional parameter per hidden state $(\phi^{(i)})$.

The likelihood of CarHMM is still compatible with the forward algorithm:
\begin{equation}
\calL_{\text{CarHMM}}(y;\Theta,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t;\Theta) \mathbf{1}_N
\label{CarHMM_likelihood}
\end{equation}
where:
%
$$P(y_t;\Theta) = \text{diag}\left(f^{(1)}(y_t|y_{t-1}), \ldots , f^{(N)}(y_t|y_{t-1}) \right)$$
%
and the first observation $y_1$ is assumed to be fixed as an initial value. The graphical model associated with the structure of a CarHMM is shown in figure (\ref{fig:CarHMM}).

@@

While both the Viterbi algorithm and the forward-backward algorithm are used in the current ecology literature, we use the \textit{forward-backward algorithm} in this work for several reasons. In particular, the forward-backward algorithm is used to find the \textit{psuedoresiduals} of a given model, which are an important tool for model validation. In addition, the forward-backward algorithm can be implemented recursively to find the probability of the fine-level states $X^*_{t^*,t}$ exactly by marginalizing out $X_t$:
%
$$P(X^*_{t^*,t} = i^*) = \sum_{i=1}^N P(X_t = i)P(X^*_{t^*,t} = i^* | X_t = i)$$
%
where $P(X_t = i)$ can be found using the forward-backward algorithm on the coarse-level Markov chain and $P(X^*_{t^*,t} = i^* | X_t = i)$ can be found by running the forward-backward algorithm on the fine-level HMM for every possible value of $i = 1, \ldots, N$.

>>>>

@@

\textit{Hidden Markov models} (or HMMs) are comprised of a sequence of $T$ unobserved states $X_t$ $(t = 1, \ldots, T)$ which forms a Markov chain and an associated sequence of $T$ (possibly high-dimensional) observations $Y_t$ $(t = 1, \ldots, T)$. These observations are often referred to as ``emissions". The index $t$ often represents equi-spaced jumps in time. In the field of animal movement, the unobserved chain usually represents the latent behaviour of an animal (e.g. foraging, resting, migrating, etc.), while the observations are often a series of step lengths and turning angles for land animals and either depth or accelerometer data (or both) for marine animals. Each random variable in the unobserved chain $X_t$ can take one of $N$ possible values, and $X$ has initial distribution $\delta \in \bbR^N$ and probability transition matrix $\Gamma \in \bbR^{N \times N}$. In particular:
%
$$\delta_i = Pr(X_1 = i); \quad i = 1,\ldots,N$$
%
$$\Gamma_{ij} = Pr(X_{t+1} = j | X_t = i); \quad t = 1, \ldots, T-1; \enspace i,j = 1,\ldots,N $$
%
We assume that $\delta$ is the stationary distribution of $\Gamma$, implying that the hidden process runs infinitely before observations are recorded. Further, the distribution of each observation $Y_t$ depends only on the value of $X_t$ and none of the preceding observations or behavioral states, i.e. $p(y_t|x_1,\ldots, x_N, y_1,\ldots,y_{t-1},y_{t+1},\ldots,y_N) = p(y_t|x_t)$. In particular, we denote the distribution of $Y_t$ given that $X_t = i$ as $f^{(i)}(y_t)$. $f^{(1)},\ldots,f^{(N)}$ are described by a vector of parameters $\Theta = (\theta^{(1)},\ldots,\theta^{(N)})$ and hereafter are referred to as ``emission distributions". A visualization of the dependence structure can be seen in (fig \ref{fig:HMM}).
In order to estimate the parameters of an HMM, suppose data $y = (y_1,\ldots,y_T)$ are observed as a realization of an unknown HMM model. The probability transition matrix $\Gamma$ and parameters of the emission distributions $\Theta$ can be estimated by maximizing the likelihood of $y$, which we denote as $\calL_{\text{HMM}}(y;\Theta,\Gamma)$, with respect to $\Gamma$ and $\Theta$. In addition, $\calL_{\text{HMM}}(y;\theta,\Gamma)$ can be calculated using the well-known \textit{forward algorithm} \cite{Zucchini:2016} as follows:
%
$$\calL_{\text{HMM}}(y;\theta,\Gamma) = \delta P(y_1;\theta) \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}_N$$
%
where $\mathbf{1}_N$ is an $N$-dimensional column vector of ones, and:
%
$$P(y_t;\theta) = \text{diag}(f^{(1)}(y_t),\ldots , f^{(N)}(y_t)).$$
%
To remove constraints when maximizing $\calL_{\text{HMM}}(y)$, we parameterize $\Gamma$ using $\eta \in \bbR^{N \times N}$ and the following link function \cite{Barajas:2017}:
%
$$\Gamma_{ij} = \frac{\exp(\eta_{ij})}{\sum_{j'=1}^N \exp(\eta_{i(j')})}, \qquad \eta_{ii} = 0 \quad i = 1, \ldots, N$$
%
$\eta_{ii}$ is set to zero for identifiability, and $\calL_{\text{HMM}}(y;\theta,\Gamma)$ can be maximized using any numerical optimizer.

@@

Hidden Markov models are useful when inferring a single unobserved process, but biological processes often involve multiple simultaneous hidden processes which can occur and at different time scales. For example, a preliminary observation of the killer whale dive data shown in (fig \ref{fig:data}) shows that the behavior of the killer whale changes between approximately hour-long periods of predominately short, shallow dives and long, deep dives. Leos-Barajas et al. \cite{Barajas:2017} encounter a similar issue when modeling the movement of a harbor porpoise in the North Sea, and use it as a motivating example when they introduce hierarchical hidden Markov models.

\begin{figure}[h!]
	\centering
	\includegraphics[width=5.5in]{../Plots/raw_data.png}
	\caption{Raw depth data of a Killer Whale off the coast of British Columbia, Canada. \textcolor{red}{I WILL REFORMAT THIS PLOT LATER.}}
	\label{fig:data}
\end{figure}

A hierarchical hidden Markov model (or HHMM) is a variation of a hidden Markov model in which $X_t$ satisfies the conditions of section \ref{subsection:HMMs}, but in addition it ``emits" yet another fine-scale hidden Markov model of length $T_t^*$. This fine-scale hidden Markov model can be either instead of \cite{Barajas:2017} or in addition to \cite{Adam:2019} a coarse-scale observation $Y_t$, which itself must satisfy the conditions from section \ref{subsection:HMMs}. The fine-scale HMM is comprised of a fine-scale Markov chain $X^*_t = (X^*_{t,1}, \ldots, X^*_{t,T^*_t})$ and fine-scale observations $Y^*_t = (Y^*_{t,1}, \ldots, Y^*_{t,T^*_t})$, each of length $T^*_t$. $X^*_t$ can take one of $N^*$ values. Note that while $N^*$ can be set to depend upon $X_t$, we assume here that it does not for simplicity in both the model presented here and the notation. Conditioned on the coarse-scale hidden state $X_t = i$, $X^*_t$ is characterized by an initial distribution $\delta^{*(i)} \in \bbR^{N^*}$ and probability transition matrix $\Gamma^{*(i)} \in \bbR^{N^* \times N^*}$:
%
$$\delta^{*(i)}_{i^*} = Pr(X^*_{t,1} = i^* | X_t = i); \quad i^* = 1,\ldots,N^*$$
%
$$\Gamma^{*(i)}_{i^*j^*} = Pr(X^*_{t,t^*+1} = j^* | X^*_{t,t^*} = i^*, X_t = i); \quad t^* = 1, \ldots, T^*_t-1; \enspace i^*,j^* = 1,\ldots,N^*$$
%
Again, although not necessary, we assume that the initial distribution $\delta^{*(i)}$ is the stationary distribution of $\Gamma^{*(i)}$. In total, there are $N$ different values that $\Gamma^{*(i)}$ and $\delta^{*(i)}$ can take- one for each possible value of the coarse-scale process $X_t$. We denote the sets of all fine-scale initial distributions and probability transition matrices as $\delta^*$ and $\Gamma^*$, respectively, where $|\delta^*| = |\Gamma^*| = N$.

Each fine-scale observation $Y^*_{t,t^*}$ depends only on the value of its corresponding fine-scale hidden state $X^*_{t,t^*}$ and coarse-scale hidden state, $X_t$. In particular, $p(y^*_{t,t^*} | y^*_{t,1}, \ldots y^*_{t,t^*-1},y^*_{t,t^*+1},\ldots,y^*_{t,T^*_t},x^*_t,x,y) = p(y^*_{t,t^*}|x^*_{t,t^*},x_t)$. We denote the distribution of $Y^*_{t,t^*}$ given that $X_t = i$ and $X^*_{t,t^*} = i^*$ as $f^{*(i,i^*)}(y^*_{t,t^*})$. Conditioned on the fact that $X_t = i$, the fine-scale distributions $f^{*(i,1)}, \ldots, f^{*(i,N^*)}$ are parameterized by an $N^*$-dimensional vector of parameters $\Theta^{*(i)} = (\theta^{*(i,1)},\ldots,\theta^{*(i,N^*)})$, where $\theta^{*(i,i^*)}$ describes the distribution $f^{*(i,i^*)}$. Depending upon the behaviour and responses being modelled, it may be desirable to force the fine-scale emission distributions to be equal across all values of $i \in \{1, \ldots, N\}: \enspace \theta^{*(1,i^*)} = \theta^{*(2,i^*)} = \ldots = \theta^{*(N,i^*)}$. A visualization of the full structure of the HHMM can be seen in (fig \ref{fig:HHMM}).

@@

One of the key assumptions of both HMMs and HHMMs is \textit{conditional independence} between observations on both the coarse and fine scale. Namely, when given the state $X_t$, $Y_t$ is assumed to be independent from all other observations, including $Y_{t-1}$. Therefore, traditional HMMs and HHMMs can fail when the observations $Y$ exhibit significant correlation in time. Examples include fluking in marine mammals in Vancouver, BC (see the results section) and the swimming behavior of horn sharks off the coast of Southern California \cite{Adam:2019}.

One way to deal with auto-correlation in fine-scale behavioral processes is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \cite{Lawler:2019}. In particular, the CarHMM describes the distribution of $(Y_t|Y_{t-1}, X_t=i)$ with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{2(i)},\phi^{(i)}\}$, where:
%
$$\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=i) = \phi^{(i)}*y_{t-1} + (1-\phi^{(i)}) * \mu^{(i)}$$
$$\mathbb{V}(Y_t| X_t = i) = \sigma^{2(i)}$$
%
The CarHMM therefore explicitly models auto-correlation into the emission distributions of the HMM while maintaining the structure needed to run the forward algorithm. It also only adds one additional parameter per possible hidden state $(\phi^{(i)})$.

The likelihood of CarHMM is still compatible with the forward algorithm, and is therefore easy to calculate:
\begin{equation}
\calL_{\text{CarHMM}}(y;\Theta,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t;\Theta) \mathbf{1}_N
\label{CarHMM_likelihood}
\end{equation}
where:
%
$$P(y_t;\Theta) = \text{diag}\left(f^{(1)}(y_t|y_{t-1}), \ldots , f^{(N)}(y_t|y_{t-1}) \right)$$
%
and the first observation $y_1$ is assumed to be fixed as an initial value. The graphical model associated with the structure of a CarHMM is shown in figure (\ref{fig:CarHMM}).

@@

While both the Viterbi algorithm and the forward-backward algorithm are used in the current ecology literature, we use the \textit{forward-backward algorithm} in this work for several reasons. First, the forward-backward algorithm is used to find the \textit{psuedoresiduals} of a given model, which are an important tool for model validation. In addition, the forward-backward algorithm can be implemented recursively to find the probability of the fine-level states $X^*_{t^*,t}$ exactly by marginalizing out $X_t$:
%
$$P(X^*_{t^*,t} = i^*) = \sum_{i=1}^N P(X_t = i)P(X^*_{t^*,t} = i^* | X_t = i)$$
%
where $P(X_t = i)$ can be found using the forward-backward algorithm on the coarse-level Markov chain and $P(X^*_{t^*,t} = i^* | X_t = i)$ can be found by running the forward-backward algorithm on the fine-scale HMM.
